---
title: "Student Performance Dashboard Recreation"
author: "Ved Thote"
date: "`r Sys.Date()`"
output: pdf_document
---

# 1. Introduction

This report analyzes the Students Performance dataset obtained from Kaggle.

**Dataset Source:**  
[Students Performance in Exams (Kaggle Dataset)](https://www.kaggle.com/datasets/spscientist/students-performance-in-exams)

The objective is to perform Exploratory Data Analysis (EDA) and create visualizations using RStudio.

# 2. Loading Required Libraries

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(corrplot)
```

# 3. Import Dataset

```{r}
data <- read.csv("StudentsPerformance.csv")
head(data)
str(data)
summary(data)
```

# 4. Data Cleaning

```{r}
colSums(is.na(data))

data$gender <- as.factor(data$gender)
data$race.ethnicity <- as.factor(data$race.ethnicity)
data$parental.level.of.education <- as.factor(data$parental.level.of.education)
data$lunch <- as.factor(data$lunch)
data$test.preparation.course <- as.factor(data$test.preparation.course)
```

# 5. Feature Engineering

```{r}
data$total_score <- data$math.score +
                    data$reading.score +
                    data$writing.score

data$average_score <- data$total_score / 3

summary(data$average_score)
```

# 6. Gender-wise Performance

```{r}
gender_avg <- data %>%
  group_by(gender) %>%
  summarise(
    Math = mean(math.score),
    Reading = mean(reading.score),
    Writing = mean(writing.score),
    Average = mean(average_score)
  )

gender_avg
```

```{r}
ggplot(data, aes(x = gender, y = average_score, fill = gender)) +
  geom_boxplot() +
  theme_minimal() +
  labs(title = "Average Score by Gender",
       x = "Gender",
       y = "Average Score")
```

**Observation:**  
Female students perform better in reading and writing, while math performance is comparable.

# 7. Test Preparation Impact

```{r}
ggplot(data, aes(x = test.preparation.course,
                 y = average_score,
                 fill = test.preparation.course)) +
  geom_boxplot() +
  theme_minimal() +
  labs(title = "Impact of Test Preparation Course",
       x = "Test Preparation Course",
       y = "Average Score")
```

**Observation:**  
Students who completed the test preparation course have higher average scores.

# 8. Parental Education Impact

```{r}
ggplot(data, aes(x = parental.level.of.education,
                 y = average_score)) +
  geom_boxplot(fill = "skyblue") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Parental Education vs Average Score",
       x = "Parental Education Level",
       y = "Average Score")
```

**Observation:**  
Higher parental education levels are associated with better student performance.

# 9. Correlation Analysis

```{r}
score_data <- data[, c("math.score",
                       "reading.score",
                       "writing.score")]

cor_matrix <- cor(score_data)
cor_matrix
```

```{r}
corrplot(cor_matrix,
         method = "color",
         addCoef.col = "black")
```

**Observation:**  
There is a strong positive correlation between all three subjects.

# 10. Score Distribution

```{r}
ggplot(data, aes(x = math.score)) +
  geom_histogram(fill = "blue", bins = 20) +
  theme_minimal() +
  labs(title = "Distribution of Math Scores")
```

```{r}
ggplot(data, aes(x = reading.score)) +
  geom_histogram(fill = "green", bins = 20) +
  theme_minimal() +
  labs(title = "Distribution of Reading Scores")
```

```{r}
ggplot(data, aes(x = writing.score)) +
  geom_histogram(fill = "purple", bins = 20) +
  theme_minimal() +
  labs(title = "Distribution of Writing Scores")
```

# 11. Grade Classification

```{r}
data$grade <- ifelse(data$average_score >= 80, "A",
                     ifelse(data$average_score >= 60, "B", "C"))

table(data$grade)
```

```{r}
ggplot(data, aes(x = grade, fill = grade)) +
  geom_bar() +
  theme_minimal() +
  labs(title = "Grade Distribution",
       x = "Grade",
       y = "Count")
```

**Observation:**  
Most students fall into Grade B category.

# 12. Conclusion

- Test preparation improves student performance.
- Strong positive correlation exists among math, reading, and writing scores.
- Parental education positively influences student outcomes.
- Female students perform better in reading and writing subjects.

